---
title: "Mastering Shiny"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
  pdf_document:
    number_sections: true
    latex_engine: pdflatex
---

[Wickham, H. (2021). Mastering shiny. O'Reilly Media, Inc.](https://mastering-shiny.org/)

```{r, eval=knitr::is_html_output(), echo=FALSE}
knitr::asis_output('[Download PDF version](mastering_shiny.pdf)')
```

# Your First Shiny App

# Basic UI

# Basic Reactivity

# Case Study: ER Injuries

# Workflow

1. Development

    1. Type *shinyapp* in .R file to insert Shiny app snippet
   
    2. Keyboard shortcut to run the app: *Ctrl+Shift+Enter*
    
    3. Relaunch app after every save with background job: 
    
       - add script `shiny-run.R` to folder with `app.R`:
       
         ```{r, eval = FALSE}
         options(shiny.autoreload = TRUE)
         shiny::runApp()
         ```
         
       - with active `shiny-run.R`, RStudio > Tools > Background Jobs > Start Background Job
       
       - copy URL from Jobs pane and run `rstudioapi::viewer("<URL>")`

2. Debugging

   1.

3. Reprexes

# Layout, Themes, HTML




