---
title: "Mastering Shiny"
output:
  pdf_document:
    number_sections: true
    latex_engine: pdflatex
---

[Wickham, H. (2021). Mastering shiny. O'Reilly Media, Inc.](https://mastering-shiny.org/)

```{r, eval=knitr::is_html_output(), echo=FALSE}
knitr::asis_output('[Download summary](mastering_shiny.pdf)')
```

# Your First Shiny App

For a basic app, create file `app.R` and add 

1. `library(shiny)` to load the shiny package
2. `ui <- fluidPage("Hello, world!")` to define the user interface (UI)
3. `server <- function(input, output, session)` to define the server function
4. `shinyApp(ui, server)` to construct and start the application

To run the app, source the document (there are other options). 

To stop the document, activate the console window and press *Esc* (there are other options). 

# Basic UI

The UI consists of inputs and outputs.

## Inputs

Insert input controls to UI by adding `{type}Input()` functions to `ui`:

- free text inputs with `textInput()`, `passwortInput()`, `textAreaInput()`
- numeric inputs with `numericInput()`, `sliderInput()`
- dates with `dateInput()`, `dateRangeInput()`
- limited choices with `selectInput()`, `radioButtons()`, `checkboxGroupInput()`
- file uploads with `fileInput()`
- action buttons with `actionButton()`, `actionLink()`

All input functions have the same first argument `inputId`. If some input function has ID `name`, than the input can be accessed in the server with `input$name`. Input functions have additional (unique) arguments to adjust their appearance.

## Outputs

Outputs in the UI create placeholders that are later filled by the server function. Like inputs, their first argument is always an ID `outputId`. If some output function has ID `plot`, than the output can be accessed in the server with `output$plot`.

Insert output placeholders to UI by adding `{type}Output()` functions to `ui`. Each `{type}Output()` function is coupled with a `render{Type}` function in `server`:

- text outputs with `textOutput()` (`renderText()`), 
- R code output with `verbatimTextOutput()` (`renderPrint()`)
- static tables with `tableOutput()` (`renderTable()`) 
- dynamic tables with `dataTableOutput()` (`renderDataTable()`)
- plots with `plotOutput()` (`renderPlot()`)

# Basic Reactivity

To connect inputs with outputs, Shiny uses a concept called *reactive expressions*.

## Reactive Expressions

Reactive expressions mean: when an input changes, all related outputs automatically update. Shiny knows when the update should be run. The code `output$greeting <- renderText(paste("hi",input$name))` informs Shiny how it could update the greeting if it needs to (e.g., if `input$name` changed). With this concept, code is no longer executed from top to bottom, but follows a graph of dependencies, which describes how inputs and outputs are connected.

## Modularity

However, Shiny updates outputs always as a whole. This can lead to undesired effects: 
```r
output$plot <- renderPlot({
  x <- rnorm(n = input$n)
  plot(x, xlim = input$range)
})
```
The random vector `x` is drawn again, when `input$range` changes. Better put the computation of `x` into a separate reactive environment. Now the value of `x` must be accessed via `x()`:
```r
x <- reactive(rnorm(n = input$n))
output$plot <- renderPlot(plot(x(), xlim = input$range))
```

If `x` should be drawn after an event (e.g., the user clicked a button), but not when `input$n` changes, use `x <- eventReactive(input$simulate, rnorm(n = input$n))`.

# Case Study: ER Injuries

Development of a richer Shiny app with the concepts seen so far. [Demo](https://hadley.shinyapps.io/ms-narrative) [Source](https://github.com/hadley/mastering-shiny/tree/master/neiss/narrative.R)

# Workflow

> One of the reasons that I’ve been able to accomplish so much is that I devote time to analysing and improving my workflow. I highly encourage you to do the same! -- Hadley Wickham

## Development

- Type *shinyapp* in .R file to insert Shiny app snippet

- Keyboard shortcut to run the app: *Ctrl+Shift+Enter*

- Relaunch app after every save with background job: 

   1. add script `shiny-run.R` to folder with `app.R`:
   
     ```{r, eval = FALSE}
     options(shiny.autoreload = TRUE)
     shiny::runApp()
     ```
     
   2. with active `shiny-run.R`, RStudio > Tools > Background Jobs > Start Background Job
   
   3. copy URL from Jobs pane and run `rstudioapi::viewer("<URL>")`

## Debugging

- Shiny automatically prints the traceback to the console

- Use interactive debugger with `browser()` in source

- Use `message()` (with `glue::glue()`) or `str()` calls to understand when a part of the code is evaluated and to show values

- Getting help: make a reprex (minimal reproducible example)

# Layout, Themes, HTML

# Graphics 

interactive graphics `plotOutput("id", click = "plot_click")` in ui makes coordinates `input$plot_click` available in server

use `req()` to avoid app action before user input

use `nearPoints(<dataset>, input$plot_click)` to get points near to the click

can also use `dblclick`, `hover`, and `brush` (together with `brushedPoints()` helper) argument

can modify a plot interactively with `reactiveVal()`

# User Feedback

# Uploads and Downloads

# Dynamic UI

# Bookmarking

# Tidy Evaluation
